---
output:
  github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Travis build status](https://travis-ci.org/daqana/dqrng.svg?branch=master)](https://travis-ci.org/daqana/dqrng)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/daqana/dqrng?branch=master&svg=true)](https://ci.appveyor.com/project/rstub/dqrng)
[![CRAN status](https://www.r-pkg.org/badges/version/dqrng)](https://cran.r-project.org/package=dqrng)
[![Coverage status](https://codecov.io/gh/daqana/dqrng/branch/master/graph/badge.svg)](https://codecov.io/github/daqana/dqrng?branch=master)
[![Downloads](http://cranlogs.r-pkg.org/badges/dqrng?color=brightgreen)](http://www.r-pkg.org/pkg/dqrng)
[![CII Best Practices](https://bestpractices.coreinfrastructure.org/projects/2157/badge)](https://bestpractices.coreinfrastructure.org/projects/2157)
[![Codacy Badge](https://api.codacy.com/project/badge/Grade/cc38be1f5900415a84010acd9ef85ce1)](https://www.codacy.com/app/rstub/dqrng?utm_source=github.com&amp;utm_medium=referral&amp;utm_content=daqana/dqrng&amp;utm_campaign=Badge_Grade)

# dqrng

The dqrng package provides fast random number generators (RNG) with good statistical properties for usage with R.
It combines these RNGs with fast distribution functions to sample from uniform, normal or exponential distributions.
Both the RNGs and the distribution functions are distributed as C++ header-only library.

## Installation

The currently released version is available from CRAN via

```r
install.packages("dqrng")
```

Intermediate releases can also be obtained via
[drat](https://cran.r-project.org/package=drat):

```r
if (!requireNamespace("drat", quietly = TRUE)) install.packages("drat")
drat::addRepo("daqana")
install.packages("dqrng")
```

## Example

Using the provided RNGs from R is deliberately similar to using R's build-in RNGs:

```{r example}
library(dqrng)
dqRNGkind("Xoroshiro128+")
dqset.seed(42)
dqrunif(5, min = 2, max = 10)
dqrexp(5, rate = 4)
```

They are quite a bit faster, though:

```{r performance}
N <- 1e7
system.time(rnorm(N))
system.time(dqrnorm(N))
```

## Feedback

All feedback (bug reports, security issues, feature requests, ...) should be provided as [issues](https://githb.com/daqana/dqrng/issues).
