---
title: "Fast Pseudo Random Number Generators for R"
author: "Ralf Stubner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fast Pseudo Random Number Generators for R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The dqrng package provides fast random number generators (RNG) with good statistical properties for usage with R.
It combines these RNGs with fast distribution functions to sample from uniform, normal or exponential distributions.
Both the RNGs and the distribution functions are distributed as C++ header-only library.

## Supported Random Number Generators

Support for the following 64bit RNGs are currently included:

* **Mersenne-Twister**  
The 64 bit variant of the well-known Mersenne-Twister, which is also used as default.
This is a conservative default that allows you to take advantage of the fast distribution functions provided by dqrng while staying close to R's default RNG (32 bit Mersenne-Twister).
* **pcg64**
The default 64 bit variant from the PCG family developed by Melissa O'Neill.
See http://www.pcg-random.org for more details.
* **Xoroshiro128+**, **Xorshift128+**, and **Xorshift1024***  
RNGs mainly developed by Sebastiano Vigna.
They are used as default RNGs in Erlang and different JavaScript engines.
See http://xoroshiro.di.unimi.it/ for more details.

Of these RNGs Xoroshiro128+ is fastest and therefore used in the examples. 

## Usage from R

Using these RNGs from R is delibratly similar to using R's build-in RNGs:

* `dqRNGkind()` is analogue to `RNGkind()` and sets the RNG
* `dqset.seed()` is analogue to `set.seed()` and sets the seed
* `dqrunif()`, `dqrnorm()`, and `dqrexp()` are analoge to `runif()`, `rnorm()`, and `rexp()` and sample from uniform, normal or exponenetial distributions

Let's look at the classical example of calculating $\pi$ via simulation. The 
basic idea is to generate a large number of random points within the unit 
square. An approximation for $\pi$ can then be calculated from the ratio of 
points within the unit circle to the total number of points. A vectorized 
implementation in R where we can switch the RNG might look like this:

```{r}
N <- 1e7
piR <- function(n, rng = runif) {
    x <- rng(n)
    y <- rng(n)
    4 * sum(sqrt(x^2 + y^2) < 1.0) / n
}
set.seed(42)
system.time(cat("pi ~= ", piR(N), "\n"))
```

Using dqrng is about three times faster:

```{r}
library(dqrng)
dqRNGkind("Xoroshiro128+")
dqset.seed(42)
system.time(cat("pi ~= ", piR(N, rng = dqrunif), "\n"))
```

Since the calculations add a constant off-set, the speed-up for the RNGs alone has to be even greater: 

```{r}
system.time(runif(N))
system.time(dqrunif(N))
```

Similar for the exponential distribution:

```{r}
system.time(rexp(N))
system.time(dqrexp(N))
```

And for the normal distribution:

```{r}
system.time(rnorm(N))
system.time(dqrnorm(N))
```

## Usage from C++

The RNGs and distributions functions can also be used from C++ at various levels of abstraction.
Technically there are three ways to make use of dqrng at the C++ level:

* use `// [[Rcpp::depends(dqrng)]]` together with `Rcpp::sourceCpp()`
* use `Rcpp::cppFunction(depends = "dqrng", ...)`
* use an R pacakge with `LinkingTo: dqrng`

Here only the first approach is shown.

### Using the compiled library functions

The functions available in R directly call corresponding C++ functions.
These functions are also available at the C++ level if you include `dqrng.h`.
Revisting the example of approximating $\pi$ we can use:

```{r, eval=FALSE, engine='Rcpp'}
// [[Rcpp::depends(dqrng)]]
#include <Rcpp.h>
#include <dqrng.h>

using Rcpp::NumericVector;
using Rcpp::sqrt;
using Rcpp::sum;
using dqrng::dqrunif;

// [[Rcpp::export]]
double piCpp(const int n) {
  dqrng::dqRNGkind("Xoroshiro128+");
  dqrng::dqset_seed(42);
  NumericVector x = dqrunif(n);
  NumericVector y = dqrunif(n);
  NumericVector d = sqrt(x*x + y*y);
  return 4.0 * sum(d < 1.0) / n;
}
/*** R
system.time(cat("pi ~= ", piCpp(1e7), "\n"))
*/
```

Note that in C++ you have to use `dqrng::dqset_seed()`, whereas the anlogue function in the R interface is called `dqrng::dqset.seed()`.

### Using the header only library

The RNG wrapper and distributions functions can be used from C++ by including `dqrng_generator.h` and `dqrng_distribution.h`.
For example, you can use the distribution functions from dqrng together with some foreign 64bit RNG.
Here a minimal [SplitMix](http://xoroshiro.di.unimi.it/splitmix64.c) generator is used together with `dqrng::normal_distribution`:

```{r, eval=FALSE, engine='Rcpp'}
#include <cstdint>
#include <Rcpp.h>
// [[Rcpp::depends(dqrng)]]
#include <dqrng_distribution.h>

struct SplitMix {
  typedef uint64_t result_type;
  static constexpr result_type max() {return UINT64_MAX;}
  static constexpr result_type min() {return 0;}
  SplitMix (result_type seed) : state(seed) {};
  inline result_type operator() () {
    result_type z = (state += 0x9e3779b97f4a7c15ULL);
    z = (z ^ (z >> 30)) * 0xbf58476d1ce4e5b9ULL;
    z = (z ^ (z >> 27)) * 0x94d049bb133111ebULL;
    return z ^ (z >> 31);
  }
  void seed(result_type seed) {state = seed;}

  private:
    result_type state;
};

// [[Rcpp::export]]
Rcpp::NumericVector splitmix_rnorm(const int n, const double mean = 0.0, const double sd = 1.0) {
  auto rng = dqrng::generator<SplitMix>(42);
  dqrng::normal_distribution dist(mean, sd);
  return dqrng::generate<dqrng::normal_distribution, Rcpp::NumericVector>(n, rng, dist);
}
/*** R
splitmix_rnorm(10)
system.time(splitmix_rnorm(1e7))
*/
```

Since SplitMix is a very fast RNG, the speed of this function is comparable to `dqrnorm`.
Alternatively, you could combine the included RNGs together with dqrng's tooling and some other distribution function.
For example, this function generates random numbers according to the normal distribution using the standard library from C++11:

```{r, eval=FALSE, engine='Rcpp'}
// [[Rcpp::plugins(cpp11)]]   
#include <random>
#include <Rcpp.h>
// [[Rcpp::depends(dqrng)]]
#include <dqrng_distribution.h>
#include <xoroshiro.hpp>

// [[Rcpp::export]]
Rcpp::NumericVector std_rnorm(const int n, const double mean = 0.0, const double sd = 1.0) {
  auto rng = dqrng::generator<xoroshiro128plus_engine>(42);
  std::normal_distribution<double> dist(mean, sd);
  return dqrng::generate<std::normal_distribution<double>, Rcpp::NumericVector>(n, rng, dist);
}
/*** R
std_rnorm(10)
system.time(std_rnorm(1e7))
*/
```

Typically this is not as fast as `dqrnorm`, but the technique is useful to support distributions not (yet) included in dqrng.
Finally you could directly use the base generators, which are provided as header-only libraries, without dqrng's tooling.
For example, the following function uses the 32 bit PCG variant together with Boost's normal distribution function:

```{r, eval=FALSE, engine='Rcpp'}
#include <Rcpp.h>
// [[Rcpp::depends(dqrng, BH)]]
#include <pcg_random.hpp>
#include <boost/random/normal_distribution.hpp>

// [[Rcpp::plugins(cpp11)]]   
// [[Rcpp::export]]
Rcpp::NumericVector boost_pcg_rnorm(const int n, const double mean = 0.0, const double sd = 1.0) {
  pcg32 rng(42);
  boost::random::normal_distribution<double> dist(mean, sd);
  Rcpp::NumericVector result(n);
  auto gen = std::bind(dist, rng);
  std::generate(result.begin(), result.end(), gen);
  return result;
}
/*** R
boost_pcg_rnorm(10)
system.time(boost_pcg_rnorm(1e7))
*/
```

This is quite fast since `boost::random::normal_distribution` uses the fast Ziggurat algorithm.

<!-- STATIC_ARBITRARY_SEED -->

