---
title: "Parallel RNG usage"
author: "Ralf Stubner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel RNG usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When you want to use random number generators (RNG) for parallel computations, you need to make sure that the sequences of random numbers used by the different processes do not overlap.
There are two main approaches to this problem:^[See for example https://www.pcg-random.org/posts/critiquing-pcg-streams.html.]

* Partition the complete sequence of random numbers produced for one seed into non-overlapping sequences and assign each process one sub-sequence.
* Re-parametrize the generator to produce independent sequences for the same seed.

The RNGs included in `dqrng` offer at least one of these methods for parallel RNG usage.
When using the R or C++ interface independent streams can be accessed using the two argument `dqset.seed(seed, stream)` or `dqset_seed(seed, stream)` functions.

# Threefry: usage from R

The Threefry engine uses internally a counter with $2^{256}$ possible states, which can be split into different substreams.
When used from R or C++ with the two argument `dqset.seed` or `dqset_seed` this counter space is split into $2^{64}$ streams with $2^{192}$ possible states each.
This is equivalent to $2^{64}$ streams with a period of $2^{194}$ each.

In the following example a matrix with random numbers is generated in parallel using the parallel package.
The resulting correlation matrix should be close to the identity matrix if the different streams are independent:

```{r, eval=FALSE}
library(parallel)
cl <- parallel::makeCluster(2)
res <- clusterApply(cl, 1:8, function(stream, seed, N) {
  library(dqrng)
  dqRNGkind("Threefry")
  dqset.seed(seed, stream)
  dqrnorm(N)
}, 42, 1e6)
stopCluster(cl)

res <- matrix(unlist(res), ncol = 8)
symnum(x = cor(res), cutpoints = c(0.001, 0.003, 0.999),
       symbols = c(" ", "?", "!", "1"),
       abbr.colnames = FALSE, corr = TRUE)
```

Correlation matrix:

    [1,] 1              
    [2,]   1            
    [3,]   ? 1          
    [4,]   ? ? 1        
    [5,] ?     ? 1      
    [6,]     ?     1    
    [7,]     ?       1  
    [8,]         ?     1
    attr(,"legend")
    [1] 0 ‘ ’ 0.001 ‘?’ 0.003 ‘!’ 0.999 ‘1’ 1

As expected the correlation matrix for the different columns is almost equal to the identity matrix.


# Xo(ro)shiro: jump ahead with OpenMP

The Xoshiro256+/++/** generators has a period of $2^{256} -1$ and offers $2^{128}$ sub-sequences that are $2^{128}$ random draws apart as well as $2^{64}$ streams that are $2^{192}$ random draws apart.
The Xoroshiro128+/++/** generators has a period of $2^{128} -1$ and offers $2^{64}$ sub-sequences that are $2^{64}$ random draws apart as well as $2^{32}$ streams that are $2^{98}$ random draws apart. 
You can go from one sub-sequence to the next using the `jump()` or `long_jump()` method and the convenience wrapper `jump(int n)` or `long_jump(int n)`, which  advances to the `n`th sub-sequence.
When used from R or C++ with the two argument `dqset.seed` and `dqset_seed` you get $2^{64}$ streams that are $2^{192}$ and $2^{64}$ random draws apart for Xoshiro256+/++/** and Xoroshiro128+/++/**, respectively.

As an example using C++ we draw and sum a large number of uniformly distributed numbers.
This is done several times sequentially as well as using OpenMP for parallelisation.
Care has been taken to keep the global RNG `rng` usable outside of the parallel block.

```{r, eval=FALSE, engine='Rcpp'}
#include <Rcpp.h>
// [[Rcpp::depends(dqrng, BH)]]
#include <xoshiro.h>
#include <dqrng_distribution.h>
// [[Rcpp::plugins(cpp11)]]
// [[Rcpp::plugins(openmp)]]
#include <omp.h>

// [[Rcpp::export]]
std::vector<double> random_sum(int n, int m) {
  dqrng::uniform_distribution dist(0.0, 1.0); // Uniform distribution [0,1)
  dqrng::xoshiro256plus rng(42);              // properly seeded rng
  std::vector<double> res(m);
  for (int i = 0; i < m; ++i) {
    double lres(0);
    for (int j = 0; j < n; ++j) {
      lres += dist(rng);
    }
    res[i] = lres / n;
  }
  return res;
}

// [[Rcpp::export]]
std::vector<double> parallel_random_sum(int n, int m, int ncores) {
  dqrng::uniform_distribution dist(0.0, 1.0); // Uniform distribution [0,1)
  dqrng::xoshiro256plusplus rng(42);          // properly seeded rng
  std::vector<double> res(m);
  // ok to use rng here
  
  #pragma omp parallel num_threads(ncores)
  {
    dqrng::xoshiro256plusplus lrng(rng);      // make thread local copy of rng 
    lrng.long_jump(omp_get_thread_num() + 1);  // advance rng by 1 ... ncores jumps

    #pragma omp for
    for (int i = 0; i < m; ++i) {
      double lres(0);
      for (int j = 0; j < n; ++j) {
        lres += dist(lrng);
      }
      res[i] = lres / n;
    }
  }
  // ok to use rng here
  return res;
}

/*** R
bm <- bench::mark(
  parallel_random_sum(1e7, 8, 4),
  random_sum(1e7, 8),
  check = FALSE
)
bm[,1:4]
*/
```


Result:

      expression                            min   median `itr/sec`
      <bch:expr>                       <bch:tm> <bch:tm>     <dbl>
    1 parallel_random_sum(1e+07, 8, 4)   98.3ms     99ms     10.1 
    2 random_sum(1e+07, 8)              270.2ms    271ms      3.68

In the previous example it was necessary to create our own RNG object. It would be nice if it were possible to use dqrng's global RNG instead. To showcase this, we implement the same procedure as before, but this time make use of `dqrng::random_64bit_accessor` to access the global RNG and its `clone(stream)` method that creates a cloned version with a different stream:

```{r, eval=FALSE, engine='Rcpp'}
#include <Rcpp.h>
// [[Rcpp::depends(dqrng, BH)]]
#include <dqrng.h>
#include <dqrng_distribution.h>
// [[Rcpp::plugins(cpp14)]]
// [[Rcpp::plugins(openmp)]]
#include <omp.h>

// [[Rcpp::export]]
std::vector<double> random_sum(int n, int m) {
  dqrng::uniform_distribution dist(0.0, 1.0); // Uniform distribution [0,1)
  auto rng = dqrng::random_64bit_accessor{};  // properly seeded rng
  std::vector<double> res(m);
  for (int i = 0; i < m; ++i) {
    double lres(0);
    for (int j = 0; j < n; ++j) {
      lres += dist(rng);
    }
    res[i] = lres / n;
  }
  return res;
}

// [[Rcpp::export]]
std::vector<double> parallel_random_sum(int n, int m, int ncores) {
  dqrng::uniform_distribution dist(0.0, 1.0); // Uniform distribution [0,1)
  auto rng = dqrng::random_64bit_accessor{};  // properly seeded rng
  std::vector<double> res(m);
  // ok to use rng here

#pragma omp parallel num_threads(ncores)
{
  // make thread local copy of rng and advance it by 1 ... ncores jumps
  auto lrng = rng.clone(omp_get_thread_num() + 1);      

#pragma omp for
  for (int i = 0; i < m; ++i) {
    double lres(0);
    for (int j = 0; j < n; ++j) {
      lres += dist(*lrng);
    }
    res[i] = lres / n;
  }
}
// ok to use rng here
return res;
}

/*** R
bm <- bench::mark(
  parallel_random_sum(1e7, 8, 4),
  random_sum(1e7, 8),
  check = FALSE
)
bm[,1:4]
*/
```

Notes:

* The thread local RNG does not make use of `Rcpp::Xptr` since it is used in parallel code. Instead a `std::unique_ptr` is used that is created using `std::make_unique`. Therefore, using this feature requires at least C++14. 
* At the moment this kind of stream handling is not supported for the Threefry engine.

# PCG: multiple streams with RcppParallel

From the PCG family we will look at pcg64, a 64-bit generator with a period of $2^{128}$.
It offers the function [`advance(int n)`](https://www.pcg-random.org/using-pcg-cpp.html#void-advance-state-type-delta), which is equivalent to `n` random draws but scales as $O(ln(n))$ instead of $O(n)$. 
In addition, it offers $2^{127}$ separate streams that can be enabled via the function [`set_stream(int n)`](https://www.pcg-random.org/using-pcg-cpp.html#void-pcg32-set-stream-state-type-stream) or the [two argument constructor](https://www.pcg-random.org/using-pcg-cpp.html#pcg32-pcg32-constructor) with `seed` and `stream`.
When used from R or C++ with the two argument `dqset.seed` and `dqset_seed` you get $2^{64}$ streams out of the possible $2^{127}$ separate streams.

In the following example a matrix with random numbers is generated in parallel using RcppParallel.
Instead of using the more traditional approach of generating the random numbers from a certain distribution, we are using the fast sampling methods from `dqrng_sample.h`.
As a consequence, we cannot use `pcg64` directly but have to wrap it as `dqrng::generator`.
The resulting correlation matrix should be close to the identity matrix if the different streams are independent:


```{r, eval=FALSE, engine='Rcpp'}
#include <Rcpp.h>
// [[Rcpp::depends(dqrng, BH)]]
#include <pcg_random.hpp>
#include <dqrng_sample.h>
// [[Rcpp::plugins(cpp11)]]
// [[Rcpp::depends(RcppParallel)]]
#include <RcppParallel.h>

struct RandomFill : public RcppParallel::Worker {
  RcppParallel::RMatrix<int> output;
  uint64_t seed;

  RandomFill(Rcpp::IntegerMatrix output, const uint64_t seed) : output(output), seed(seed) {};

  void operator()(std::size_t begin, std::size_t end) {
    auto rng = dqrng::generator<pcg64>(seed, end);
    for (std::size_t col = begin; col < end; ++col) {
      auto sampled = dqrng::sample::sample<std::vector<int>, uint32_t>(*rng, 100000, output.nrow(), true);
      RcppParallel::RMatrix<int>::Column column = output.column(col);
      std::copy(sampled.begin(), sampled.end(), column.begin());
    }
  }
};

// [[Rcpp::export]]
Rcpp::IntegerMatrix parallel_random_matrix(const int n, const int m, const int ncores) {
  Rcpp::IntegerMatrix res(n, m);
  RandomFill randomFill(res, 42);
  RcppParallel::parallelFor(0, m, randomFill, m/ncores + 1);
  return res;
}

/*** R
res <- parallel_random_matrix(1e6, 8, 4)
head(res)
symnum(x = cor(res), cutpoints = c(0.001, 0.003, 0.999),
       symbols = c(" ", "?", "!", "1"),
       abbr.colnames = FALSE, corr = TRUE)
*/
```

Head of the random matrix:

          [,1]  [,2]  [,3]  [,4]  [,5]  [,6]  [,7]  [,8]
    [1,] 67984 16279 69262  7126 21441 37720 51107 51045
    [2,] 69310 21713 82885 81157 54051  5261 91165 17833
    [3,] 76742 31232 78953  4626 94939 29416 85652 78296
    [4,] 76349 47427  1770 37957 33888 59134 94591 65793
    [5,] 85008 89224 43493  7925 60866  2464 14080 10763
    [6,] 38017 88509 51195 73086  1883 68193 75259 62216

Correlation matrix:

    [1,] 1              
    [2,]   1            
    [3,]   ? 1          
    [4,]     ? 1        
    [5,]         1      
    [6,] ? ?     ? 1    
    [7,]     ?       1  
    [8,]     ?         1
    attr(,"legend")
    [1] 0 ‘ ’ 0.001 ‘?’ 0.003 ‘!’ 0.999 ‘1’ 1

So as expected the correlation matrix is almost equal to the identity matrix.
